Amendments Summary

### 2024-12-25 Thesis/01-chap1.qmd
update the final thesis

@buja2009statistical introduced a visual inference framework that formalised a hypothesis testing of graphical representations of data (henceforth referred to as the _data plot_) via the lineup protocol. The protocol is inspired by the police lineup technique employed in eyewitness identification of criminal suspects. Briefly, the protocol comprises $m$ randomly positioned plots, where one position presents the data plot, while the remaining $m-1$ plots present the plots with the same graphical structure, except that the data has been replaced with data consistent with the null hypothesis $H_0$ (henceforth referred to as _null plots_). To compute the $p$-value of the visual test, the lineup will be independently presented to a number of participants, asking them to pick the most different plot. Under $H_0$, the data plot is expected to be indistinguishable from the null plots, and the probability of correctly identifying the data plot by an observer is $1/m$. If a large number of participants correctly identify the data plot, the corresponding $p$-value will be small, indicating strong evidence against $H_0$. This protocol provides a calibration of the data plot against the null plots, ensuring that the data plot is not over-interpreted.

@@

Regression diagnostics is a well-established field with extensive literature addressing various types of regression models, such as generalized linear models [@hardin2007generalized], mixed-effects models [@pinheiro2006mixed], panel data models [@baltagi2008econometric], and survival models [@baltagi2008econometric]. This thesis, however, focuses on the classical normal linear regression model. Further discussions on extending the methods established in this thesis to other types of regression models are provided in @sec-second-paper-limitation.

>>>>

@buja2009statistical introduced a visual inference framework that formalised a hypothesis testing of graphical representations of data (henceforth referred to as the _data plot_) via the lineup protocol (see @fig-introduction-lineup for an example). The protocol is inspired by the police lineup technique employed in eyewitness identification of criminal suspects. Briefly, the protocol comprises $m$ randomly positioned plots, where one position presents the data plot, while the remaining $m-1$ plots present the plots with the same graphical structure, except that the data has been replaced with data consistent with the null hypothesis $H_0$ (henceforth referred to as _null plots_). To compute the $p$-value of the visual test, the lineup will be independently presented to a number of participants, asking them to pick the most different plot. Under $H_0$, the data plot is expected to be indistinguishable from the null plots, and the probability of correctly identifying the data plot by an observer is $1/m$. If a large number of participants correctly identify the data plot, the corresponding $p$-value will be small, indicating strong evidence against $H_0$. This protocol provides a calibration of the data plot against the null plots, ensuring that the data plot is not over-interpreted.

@@

Regression diagnostics is a well-established field with extensive literature, and a more detailed discussion will be provided in Chapter 2. This field also encompasses a variety of regression models, including generalized linear models [@hardin2007generalized], mixed-effects models [@pinheiro2006mixed], panel data models [@baltagi2008econometric], and survival models [@baltagi2008econometric]. This thesis, however, focuses on the classical normal linear regression model. Further discussions on extending the methods established in this thesis to other types of regression models are provided in @sec-second-paper-limitation.

```{r fig-introduction-lineup, fig.pos="t!", fig.height = 7, fig.width = 7, out.width = "100%", fig.cap = "A lineup is used to conduct visual testing, as demonstrated in this example. The observed data’s residual plot is placed among 19 null plots generated from a standard error model. Human judges are then asked to examine the lineup and identify the plot they find most distinct. The $p$-value is calculated based on how often the judges correctly identify the data plot, which is located at position $7$ and exhibits heteroskedasticity. A small $p$-value indicates a substantial agreement among the judges in selecting the data plot."}
set.seed(10086)
visage::heter_model(a = 0, b = 60)$gen_lineup(300) |>
  visage::VI_MODEL$plot_lineup(theme = ggplot2::theme_light(), remove_axis = TRUE, remove_grid_line = TRUE)
```

### 2024-12-25 Thesis/abstract.qmd
upload the original copy
update the final thesis

This work is motivated by the need for an automated approach to diagnose regression models through residual plots with reliability and consistency. While numerical hypothesis tests are commonly used, residual plots remain essential in regression diagnostics because conventional tests are limited to specific types of model departures and tend to be overly sensitive. Visual inference using the lineup protocol offers a less sensitive and more broadly applicable alternative, yet its dependence on human judgment limits scalability. This research addresses these limitations by automating the assessment process.

This research presents three original contributions. The first contribution provides evidence for the effectiveness of visual inference in regression diagnostics through a human subject experiment, demonstrating the benefits of using the lineup protocol for reliable and consistent reading of residual plots. The second contribution introduces a computer vision model to automate the assessment of residual plots, addressing the scalability limitation of the lineup protocol. The third contribution presents an R package and Shiny app, providing a user-friendly interface for analysts to leverage the computer vision model and supporting tools for diagnostic purposes. These contributions advance the field of artificial intelligence for data visualization, enabling more efficient and accurate regression diagnostics.

@@

**Date**: 26th August 2024

>>>>

This work is motivated by the need for an automated approach to diagnose regression models through residual plots with reliability and consistency. While numerical hypothesis tests are commonly used, residual plots remain essential in regression diagnostics because conventional tests are limited to specific types of model departures and tend to be overly sensitive. This research builds on a method called the lineup protocol, a statistical visualization approach that embeds a data plot among null plots to compare patterns arising by chance with those observed in the data. This process is part of visual inference, where human judgment is used to make statistical inferences from data visualizations. However, the reliance on human judgment in visual inference poses challenges for scalability and consistency.

To overcome these challenges, this research presents three original contributions. The first contribution provides evidence for the effectiveness of visual inference in regression diagnostics through a human subject experiment, demonstrating the benefits of using the lineup protocol for reliable and consistent reading of residual plots. The second contribution introduces a computer vision model to automate the assessment of residual plots, addressing the scalability limitation of the lineup protocol. The third contribution presents an R package and Shiny app, providing a user-friendly interface for analysts to leverage the computer vision model and supporting tools for diagnostic purposes. These contributions advance the field of artificial intelligence for data visualization, enabling more efficient and accurate regression diagnostics.

@@

**Date**: 4th September 2024
